Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-18 appendix-EPQ.tex
add files

\EPQisClosedUnderProjection*

@@

\EPQasMultiInMRpriforSequence*

@@

\EPQueueisStepByStepLinearizability*

@@

@@

@@

>>>>

\begin{restatable}{lemma}{EPQisClosedUnderProjection}
\label{lemma:EPQ is closed under projection}
$\textit{EPQ}$ is closed under projection.
\end{restatable}

@@

%\EPQasMultiInMRpriforSequence*

\begin{restatable}{lemma}{EPQasMultiInMRpriforSequence}
\label{lemma:EPQ as multi in MRpri for sequence}
Given a data-differentiated sequential execution $e$, $e \in \textit{EPQ}$, if and only if, $\forall e' \in \textit{proj}(e)$ and $\forall R \in \textit{last}(e')$, we have $e' \in \textit{MS}(R)$.
\end{restatable}

@@

%\EPQueueisStepByStepLinearizability*

\begin{restatable}{lemma}{EPQueueisStepByStepLinearizability}
\label{lemma:EPQ is step-by-step linearizability}
$\textit{PQ}$ is step-by-step linearizability.
\end{restatable}

@@

To facilitate our proof of latter sections, we need to separate $\textit{EPQ}_1$ into two cases: (1) no matched pair of $\textit{put}$ and $\textit{rm}$ in $u \cdot v \cdot w$ has priority $\textit{pri}$, (2) some matched pair of $\textit{put}$ and $\textit{rm}$ in $u \cdot v \cdot w$ has priority $\textit{pri}$.  Therefore, we separate $\textit{EPQ}_1$ into two cases:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $\textit{EPQ}_1^{>}(u,v,w,\textit{pri}) \equiv
(\textit{noRE}(u \cdot v \cdot w)) \wedge
(\textit{LI}(\textit{pri}, u \cdot v \cdot w)) \wedge
(\textit{LI-U}(\textit{pri},u \cdot v \cdot w)) \wedge
(\textit{matched-C}(u \cdot v,\textit{pri}) )$

\item[-] $\textit{EPQ}_1^{=}(u,v,w,\textit{pri}) \equiv
(\textit{noRE}(u \cdot v \cdot w)) \wedge
(\textit{LE'I}(\textit{pri}, u \cdot v \cdot w)) \wedge
(\textit{LI-U}(\textit{pri},u \cdot v \cdot w)) \wedge
(\textit{matched-C}(u \cdot v,\textit{pri}) ) \wedge
(\textit{putInSeq}(u \cdot v \cdot w,u,\textit{pri}))$
\end{itemize}

For $\textit{EPQ}_2$, we also need to distinguish two cases: (1) no matched pair of $\textit{put}$ and $\textit{rm}$ in $u \cdot v$ has priority $\textit{pri}$, (2) some matched pair of $\textit{put}$ and $\textit{rm}$ in $u \cdot v$ has priority $\textit{pri}$. Therefore, we separate $\textit{EPQ}_2$ into two rules:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $\textit{EPQ}_2^{>}(u,v) \equiv
(\textit{noRE}(u \cdot v)) \wedge
(\textit{LEI}(\textit{pri},u \cdot v)) \wedge
(\textit{LI-M}(\textit{pri},u \cdot v)) \wedge
(\textit{putInSeq}(u \cdot v,u,\textit{pri}))$.

\item[-] $\textit{EPQ}_2^{=}(u,v) \equiv
(\textit{noRE}(u \cdot v)) \wedge
(\textit{LEI}(\textit{pri},u \cdot v)) \wedge
(\textit{LE'I-M}(\textit{pri},u \cdot v)) \wedge
(\textit{putInSeq}(u \cdot v,u,\textit{pri}))$.
\end{itemize}

@@

Before we go to investigate co-regular of each rules, we use the results in \cite{Bouajjani:2015} to simplify our work. \cite{Bouajjani:2015} states that checking linearizability w.r.t queue can be reduced into checking emptiness of intersection between $\mathcal{I}$ and a set of automata. Given a data-differentiated execution $e$, let $e \vert_{i}$ be an execution generated from $e$ by erasing call and return actions of items that does not use priority $i$ (does not influence $\textit{rm}(\textit{empty})$). We call a extended priority queue execution with only one priority a single-priority execution. Let $\textit{transToQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. We can see that for each $e \in \textit{EPQ}$ and each priority $i$, $\textit{transToQueue}(e \vert_{i})$ satisfy FIFO (first in first out) property.

Given an execution of queue, we say that it is differentiated \cite{Wolper:1986}, if each item is enqueued at most once. \cite{Bouajjani:2015} states that, given a differentiated queue execution $e$ without $\textit{deq}(\textit{empty})$, $e$ is not linearizable with respect to queue, if one of the following cases holds for some $a,b$: (1) $\textit{deq}(b) <_{hb} \textit{enq}(b)$, (2) there are no $\textit{enq}(b)$ and at least one $\textit{deq}(b)$, (3) there are one $\textit{enq}(b)$ and more than one $\textit{deq}(b)$, and (4) $\textit{enq}(a) <_{\textit{hb}} \textit{enq}(b)$, and $\textit{deq}(b) <_{\textit{hb}} \textit{deq}(a)$, or $\textit{deq}(a)$ does not exists. For each such case, we can construct a witness automata for extended priority queue.

@@

The notion of left-right constraint used for extended priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}. Let us introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it correspond to the existence of the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Given a data-differentiated $\textit{pri}_x$-execution $e$ without $\textit{rm}(\textit{empty})$. Let $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ be the only  method events of $e$ with priority $\textit{pri}_x$. The left-right constraint of $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ is the graph $G$ where:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] the nodes are the items of $e$, to which we add a node,

\item[-] there is an edge from item $d_1$ to $x$, if $\textit{put}(d_1,\_)$ happens before $\textit{put}(x,\textit{pri}_x)$ or $\textit{rm}(x)$,

\item[-] there is an edge from $x$ to item $d_1$, if $\textit{rm}(x)$ happens before $\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists in $h$,

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item with maximal priority (for example, $x$) in $G$, we say that $x$ is covered by $d_1,\ldots,d_m$. To state the effectiveness of left-right constraint, take the execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} as an example, we can see that $b$ is covered by $f,e,d$. We need to prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$, as stated by the following lemma:

\begin{restatable}{lemma}{LinEqualsConstraintforEPQOneLar}
\label{lemma:Lin Equals Constraint for EPQ1Lar}
Given a data-differentiated $\textit{pri}_x$-execution $e$ with $\textit{last}(e) = \textit{EPQ}_1^{>}$. Let $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ be method events of $e$ with maximal priority. Let $G$ be the graph representing the left-right constraint of $\textit{put}(x)$ and $\textit{rm}(x)$. $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$, if and only if $G$ has no cycle going through $x$.
\end{restatable}
